---
title: "GitHubRintro"
author: "Libby Mojica"
date: "2023-01-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

#Course information
<https://github.com/Data-Sci-Intro-2023/Week-1-Intro/blob/main/lessons/intro-basics.md>
Video: <https://www.youtube.com/watch?v=l3-xEn1KQxY&t=929s>

```{r}
source("setup.R")
```

# 6.1 Load penguins dataset

```{r}
data("penguins")

penguins
```

# 6.2 Learning data types in R

vector - character, 1 dimensional list - 1 dimensional matrix - 2
dimensional

```{r}
class(penguins)
str(penguins)
```

```{r}
class(penguins$species)
```

Data type: Lists

```{r}
myList <- list("apple", 1993, FALSE, penguins)
str(myList)
```

```{r}
names(myList) <- c("fruit", "year", "logic", "data")
names(myList)
```

# 6.3 Learning Indexing

for lists we use double brackets pulls first thing in list

```{r}
myList[[1]]
```

index vectors, single bracket for brackets pulls second item in list

```{r index vector}

myVector <- c("apple", "banana", "pear")
myVector[2]
```

Index data frames examples

```{r}
# dataframe[row(s), then columns()]
# code will pull rows 1 to 5 on column 2
penguins[1:5, 2]

penguins[1:5, "island"]

penguins[1, 1:5]

penguins[1:5, c("species","sex")]

penguins[penguins$sex=='female',]



```

## 6.3.1 Exercises

1.  Why don't the following lines of code work? Tweak each one so the
    code runs

```{r}
myList["fruit"]
penguins$flipper_length_mm
penguins[penguins$island=='Dream',]
```

2.  How many species are in the penguins dataset? What islands were the
    data collected for? (Note: the unique() function might help)

```{r}
#There are 3 penguin species
unique(penguins$species)
```

3.Use indexing to create a new data frame that has only 3 columns:
species, island and flipper length columns, and subset all rows for just
the 'Dream' island.

```{r}
newpenguins <- penguins[penguins$island=='Dream', c("species", "island", "flipper_length_mm")]
newpenguins
```

4.Use indexing and the mean() function to find the average flipper
length for the Adelie species on Dream island. (Note: explore the mean()
function and how to deal with NA values).

```{r}
penguins %>%
  filter(species == "Adelie", island == "Dream")%>% 
  summarise (mean = mean(penguins$flipper_length_mm, na.rm = TRUE))

```

# 6.4 Learning the dplyr package

Subset rows with filter() Select columns with select() Create new
variables with mutate() Converts to to a grouped table group_by() Pipes
express a sequence of operations

## 6.3.2 Exercises

1.Reorder the variables in penguins so that year is the first column
followed by the rest (Hint: look into the use of everything()).

```{r year column first}
penguins [,c(8,1,2,3,4,5,6,7)]
```

2.Create a new column called 'size_group' where individuals with body
mass greater than the overall average are called 'large' and those
smaller are called 'small'.

```{r create size group}
penguins %>% 
  mutate(size_group = if_else(body_mass_g < mean(body_mass_g, na.rm = TRUE), 
                              paste0("small"), paste0("large") ))
```

3.  Find out which year for each species were individuals on average the
    largest according to body mass.

```{r}
penguins%>%
  group_by(year, species) %>%
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE)) %>%
  arrange(species, desc(mean_mass))

```

4.  You want to filter data for years that are not in a vector of given
    years, but this code doesn't work. Tweak it so that it does.

```{r}
penguins %>% 
  filter(year != c(2008, 2009))
```

## 7.1 Exercises

1.  Make a barplot showing the average flipper length for each species.

```{r}
penguins %>%
  ggplot +
  geom_bar( mapping = aes(x = flipper_length_mm, color = species))
```

2.  Make a scatter plot of bill length compared to bill depth but only
    for observations on the Dream island.

```{r}
penguins %>%
  filter(island == "Dream") %>% 
  ggplot +
  geom_point(mapping = aes(bill_length_mm, bill_depth_mm, color = species))
```
