---
title: "Week1, Day 2 Spatial Data"
author: "Libby Mojica"
date: "2023-01-20"
output:
  html_document: default
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
---

#```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
#```

# Intro to Spatial Data in R

Video <https://www.youtube.com/watch?v=T3M40tbbFwg> lesson
<https://github.com/Data-Sci-Intro-2023/Week-1-Intro/blob/main/lessons/intro-spatial.md>

# 2. Import and manipulate spatial data

```{r setup}
source("setup.R")
```

## 2.1 Vector Data with tigris and tmap

### tigris

import Colorado counties with tigris

```{r}
counties <- counties(state = "CO")
```

import roads in Larimer county limit what data we call because large
amount of data

```{r}

roads <- roads(state = "CO", county = "Larimer")
```

### tmap

Set tmap mode to interactive viewing

```{r}
tmap_mode("view")
```

Quick thematic map

```{r}
qtm(counties)
```

We can click on polygons on map in viewer to see details. For a more
detailed map use tm_shape instead of qtm. It's a better method to create
customized maps.

```{r}
tm_shape(counties)+ 
  tm_polygons()
```

Quick thematic map:

```{r}
qtm(counties)+
  qtm(roads)
```

Look at the class of counties

```{r}
class(counties)
```

### sf

sf (simple features) is a package to work with vector data, adds a
geometry column

Filter road data set for only Poudre Canyon Hwy

```{r}
poudre_hwy <- roads %>% 
  filter(FULLNAME == "Poudre Canyon Hwy")
qtm(poudre_hwy)
```

### Points

When we create poudre_points object it has no spatial information (if
you look at class(poudre_points) it returns "data.frame")

```{r}
poudre_points <- data.frame(name = c("Mishawaka", "Rustic", "Blue Lake Trailhead"),
                            long = c(-105.35634, -105.58159, -105.85563),
                            lat = c(40.68752, 40.69687, 40.57960))
```

Convert to spatial coordinate reference system WGS 1984 is crs = 4326

```{r}
poudre_points_sf <- st_as_sf(poudre_points, coords = c("long", "lat"), crs = 4326)
poudre_points_sf
```

View on map

```{r}
qtm(poudre_hwy)+
  qtm(poudre_points_sf)
```

## 2.2 Raster data

elevatr package, imports raster dem z specifies the zoom level, or
resolution of the raster, z = 7 is 1 km resolution

```{r}
elevation <- get_elev_raster(counties, z = 7)
elevation
qtm(elevation)
```

change elevation display continuous palette for map symbol add more
informative legend title looks more gradual symbols because continuous
elevation

```{r}
tm_shape(elevation)+
  tm_raster(style = "cont", title = "Elevation (m)")
```

### terra package

Raster data management package, but only if we convert to a terra raster
object Converts data from "RasterLayer" to "SpatRaster"

```{r}
# Now data in terra format as a SpatRaster we can use terra functions
elevation <- rast(elevation)
elevation
```

Update the name of the file

```{r}
names(elevation) <- "Elevation"
elevation
```

### Coordinate Reference Systems

Check projection with st_crs function we find this data set in NAD83

```{r}
st_crs(counties)
```

Check if two data sets are in same projection returns false, so need to
reproject one of the datasets

```{r}
crs(counties) == crs(elevation)
```

To reproject elevation layer to match counties

```{r}
elevation_prj <- terra::project(elevation, counties)
```

Now that both datasets in same projection we can crop elevation to match
the county boundary 
then crop elevation to Colorado counties extent

```{r}
elevation_crop <- crop(elevation, ext(counties))
qtm(elevation_crop)
```

# 3.Read and write spatial data
## 3.1 Writing spatial data
Save sf or vector data

```{r}
# write_sf(counties, append = FALSE, "data/counties.shp")
write_sf(poudre_hwy, append = FALSE,  "data/poudre_hwy.shp")
write_sf(poudre_points_sf, append = FALSE, "data/poudre_points.shp")
```

Save raster data

```{r}
writeRaster(elevation_crop,  overwrite=TRUE, "data/elevation_larimer.tif")
```

## 3.2 Save RData files

```{r}
save(poudre_hwy, poudre_points_sf, file = "data/spatial_objects.RData")
```

Test removing and reloading spatial objects saved as RData file

```{r}
#rm(poudre_hwy, poudre_points_sf)
#load("data/spatial_objects.RData")
```

# 4. Exercises

Filter out the counties data set to only include counties:
Larimer, Denver, and Pueblo.
```{r}
counties %>%
 filter(NAME == 'Larimer'|NAME =='Denver'|NAME =='Pueblo')
```

Make a map of the counties data colored by county area. Make a second map of counties colored by their total area of water.
```{r}
#Combine area land and area water to make total county area field
#then map total county area
counties %>%
  mutate(Area_total = ALAND + AWATER) %>% 
  ggplot(.) +
  geom_sf(aes(fill = Area_total))
```

```{r}
#Map counties by total water area
counties %>%
    ggplot(.) +
  geom_sf(aes(fill = AWATER))
```


Make a barplot comparing the elevation of your 3 points in the Poudre Canyon (note: explore the extract() function in the terra package).
```{r}
#create a new column that adds the elevation for each point
poudre_points_sf$pt_elev <- terra::extract(elevation, vect(poudre_points_sf))[,2]
```

Then plot elevation on a barplot
```{r}
poudre_points_sf %>% 
  ggplot() +
  geom_col(mapping = aes(x = name, y = pt_elev)) + 
  ylab("Elevation") + xlab("Location")
```

Why are there 4 features in our Poudre Canyon Highway variable instead of 1?
```{r}
#There are 4 segments of the road saved in the roads data set. You can see them plotted by column LINEARID
tm_shape(poudre_hwy)+
  tm_lines(col = "LINEARID")

```

